Differential invariant signatures for planar Lie group transformations of images

### 2020-11-16 paper/newpaper.tex

\section{Introduction}
\todo[inline]{
Richard: 
Start on invariant tensor theorem
Generate new pictures, add function details
SE(2) jobs inline
Mobius, Conformal -- finish code, tidy writing
First bash at experimental section

Stephen:
Rearrange examples text to IVT section, write about signatures here
Fix up tables, check bib
Examples for SA(2), A(2) -- need to think about sig here
Moving frame intro
Projective -- tidy writing, especially the names of the invariants
Colour
First bash at conclusion

Later:
Introduction
Count of derivatives and what to say
}

\todo[inline]{Examples to be numbered like theorems}
\todo[inline]{This is a start. To be completed later.}

@@

An alternative option is to seek $\mathcal{G}$-invariant functions on $M$, which is what we consider here. In this case, rather than considering the image of the object, one constructs an invariant signature of some kind, which represents the object modulo the action of elements from the relevant group. This has been considered extensively by Olver and co-authors for the case of outlines of objects~\cite{Calabi1998,Hoff2013}. In this paper we consider the case of signatures for images rather than curve outlines.

@@

@@

In order to define differential invariants, it is necessary to identify how the group transformation $\varphi$ acts on derivatives of the function $f$. This is known as the jet space, or jet bundle~\cite{OlverEIS}. The $d$-th order jet space $J^d$ consists of the function itself and the set of all partial derivatives of the function up to order $d$. (Note that there are $\left( \begin{array}{c} d+k-1 \\ k \end{array} \right)$ partial derivatives of order $d$). As the group $\mathcal{G}$ is local, it acts naturally on derivatives of the transformation, and so each individual transformation $\varphi \in \mathcal{G}$ can be prolonged to act on the jet space $J^d$. We can thus prolong the action of the group and consider the action of $\mathcal{G}$ on not just $f$, but all derivatives of $f$ up to order $d$. A $d$-th order differential invariant is then a local scalar function $I: J^{d} (f,\Omega, k) \to \mathbb{R}$ that is invariant under the action of $\mathcal{G}$, i.e., $I(\varphi(f,\Omega, k)) = I(f, \Omega, k) \forall \varphi \in \mathcal{G}$.

An $m$-dimensional signature set is the image of $m$ invariants $\mathcal{I}=(I_1,I_2,\dots,I_m)$, i.e., the set $\mathcal{I}(f)(\Omega)\subset \mathbb{R}^m$.  A signature set is complete for $f$ if it determines $f$ up to transformations from $\mathcal{G}$, i.e., if $\mathcal{I}(\tilde f)(\tilde\Omega) = \mathcal{I}(f)(\Omega)$ implies that there exists a transformation $\varphi\in \mathcal{G}$ such that $\tilde f = \varphi\cdot f$ and $\tilde \Omega = \varphi(\Omega)$.

@@

An algorithm to compute invariants and determine complete signature sets is
described in \cite{Olver2005}. First, the entire set of invariants is
computed using the method of moving frames. Second, determine the
differential invariant order $s$ of $f$ \todo[inline]{we need to define the differential invariant order}. Then the signature set of a maximal
set of functionally independent invariants of order $s+1$ is complete. A
third step is possible, in which the dimension of the signature set is
reduced by eliminating those invariants whose values can be computed from the
values of derivatives of some subset of the invariants.

@@

Under this formula we end with a factor $(\det A)^{-r}$ in place of $(\det
A)^{-1}$ that we saw previously, when all $n$ copies of $\mathbb{R}^2$ are
transformed by the same affine function $\mathbf{x} \mapsto A\mathbf{x} +
\mathbf{b}$. $r$ is referred to as the weight of the partial
transvectant. It can also be useful to consider the maximum order of
derivative in the transvectant, which we term degree, e.g., $f_{xy}^2 f_x
f_y$ has degree 2. Note also that the individual pairwise Omega processes commute.

@@

Hence, we can form a signature for $SA(2)$ by choosing any combination of
partial transvectants. We'll choose a signature with three components so that
it can be visualised as a three-dimensional surface; note that this will not
be complete and may well have a very large bad set. For example, we can choose the function value together with the first two partial transvectants to form the signature:
\begin{equation*}
    \left(f, f_{xx}f_{yy}-f_{xy}^2, f_x^2f_{yy} - 2f_xf_yf_{xy} + f_{xx}f_y^2\right).
\end{equation*}

\subsubsection{Example}
Consider as our image, the function $f(x, y) = 0.5 - 0.2x - 0.3y - 0.05x^2 +
0.03xy + 0.04y^2 + 0.03y^3 +0.001x^3 - 0.0015x^2y + 0.002xy^2 - 0.0005y^3$
defined on $[0, 1]\times[0, 1]$.
\todo[inline]{remake figure with modified matrix. Also modify function so it has some features like a critical point}
We'll subject our image to a special affine transformation:
\begin{equation*}
\mathbf{x} \mapsto \frac{1}{1.1}\begin{bmatrix} 0.6875 & 1.1 \\ -0.6 & 0.8 \end{bmatrix}\mathbf{x} + \begin{bmatrix} 0.1 \\ -0.2 \end{bmatrix}
\end{equation*}

@@

\subsection{The Euclidean Group E(2)}
An element of the Euclidean group $E(2)$ maps an element $(x, y)$ of
$\mathbb{R}^2$ to $(\bar{x}, \bar{y})$ as 
\begin{equation*}
  \begin{aligned}
    \bar{x} &=  x\varepsilon\cos\theta  - y\varepsilon\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
where $\varepsilon \in \{-1, 1\}$ and the remaining parameters are the same
as for $SE(2)$. If $\varepsilon = 1$ it is a rigid transformation, and if
$\varepsilon = -1$ it contains a reflection.

>>>>

\section{Introduction}
%\todo[inline]{
%Richard: 
%Start on invariant tensor theorem
%Generate new pictures, add function details
%SE(2) jobs inline
%Mobius, Conformal -- finish code, tidy writing
%First bash at experimental section

%Stephen:
%Rearrange examples text to IVT section, write about signatures here
%Fix up tables, check bib
%Examples for SA(2), A(2) -- need to think about sig here
%Moving frame intro
%Projective -- tidy writing, especially the names of the invariants
%Colour
%First bash at conclusion

%Later:
%Introduction
%Count of derivatives and what to say
%}

%\todo[inline]{Examples to be numbered like theorems}
%\todo[inline]{This is a start. To be completed later.}

@@

An alternative option is to seek $\mathcal{G}$-invariant functions on $M$, which is what we consider here. In this case, rather than considering the image of the object, one constructs an invariant signature of some kind, which represents the object modulo the action of elements from the relevant group. This has been considered extensively by Olver and co-authors for the case of outlines of objects~\cite{Calabi1998,Hoff2013}. In this paper we consider the case of signatures for images rather than outlines of curves.

@@

In this paper we investigate differential invariant signatures for images and parts of images under planar spatial transformations, for both grayscale and colour images. The remainder of this introductory sectin provides a description of the notation that we use in this paper, followed by a summary of relevant work on invariants. We also introduce the way that we will present examples of the signatures for the various groups we consider. Following this, in the next three sections we describe three distinct methods of computing differential invariants: by tensor contraction, using the theory of transvectants, and via the method of moving frames, applying the methods to a variety of the planar Lie groups. We consider the difference that colour images make in Section~\ref{sec:colour}, and then discuss the number of derivatives required and the effect this has on the practicality of the method.

@@

In order to define differential invariants, it is necessary to identify how the group transformation $\varphi$ acts on derivatives of the function $f$ via the jet space~\cite{OlverEIS}. The $d$-th order jet space $J^d$ consists of the function itself and the set of all partial derivatives of the function up to order $d$. (Note that there are $\left( \begin{array}{c} d+k-1 \\ k \end{array} \right)$ partial derivatives of order $d$.) As the group $\mathcal{G}$ is local, it acts naturally on derivatives of the transformation, and so each individual transformation $\varphi \in \mathcal{G}$ can be prolonged to act on the jet space $J^d$. We can thus prolong the action of the group and consider the action of $\mathcal{G}$ on not just $f$, but all derivatives of $f$ up to order $d$. A $d$-th order differential invariant is then a local scalar function $I: J^{d} (f,\Omega, k) \to \mathbb{R}$ that is invariant under the action of $\mathcal{G}$, i.e., $I(\varphi(f,\Omega, k)) = I(f, \Omega, k) \forall \varphi \in \mathcal{G}$. It can also be useful to consider the maximum order of derivative in the invariant, which we term degree, e.g., $f_{xy}^2 f_x f_y$ has degree 2.

An $m$-dimensional signature set is the image of $m$ invariants $\mathcal{I}=(I_1,I_2,\dots,I_m)$, i.e., the set $\mathcal{I}(f)(\Omega)\subset \mathbb{R}^m$.  A signature set is complete for $f$ if it determines $f$ up to transformations from $\mathcal{G}$, i.e., if $\mathcal{I}(\tilde f)(\tilde\Omega) = \mathcal{I}(f)(\Omega)$ implies that there exists a transformation $\varphi\in \mathcal{G}$ such that $\tilde f = \varphi\cdot f$ and $\tilde \Omega = \varphi(\Omega)$. If the signature is not complete, then there will be a \emph{bad set} of images that cannot be differentiated between. ***

@@

An algorithm to compute invariants and determine complete signature sets is
described in \cite{Olver2005}. First, the entire set of invariants is
computed using the method of moving frames. Second, determine the
differential invariant order $s$ of $f$. Then the signature set of a maximal
set of functionally independent invariants of order $s+1$ is complete. A
third step is possible, in which the dimension of the signature set is
reduced by eliminating those invariants whose values can be computed from the
values of derivatives of some subset of the invariants.

@@

Under this formula we end with a factor $(\det A)^{-r}$ in place of $(\det
A)^{-1}$ that we saw previously, when all $n$ copies of $\mathbb{R}^2$ are
transformed by the same affine function $\mathbf{x} \mapsto A\mathbf{x} +
\mathbf{b}$. $r$ is referred to as the weight of the partial
transvectant. Note that the individual pairwise Omega processes commute.

@@

Hence, we can form a signature for $SA(2)$ by choosing any combination of
partial transvectants. We choose a signature with three components so that
it can be visualised as a three-dimensional surface; note that this will not
be complete and may well have a very large bad set. For example, we can choose the function value together with the first two partial transvectants to form the signature:
\begin{equation*}
    \left(f, f_{xx}f_{yy}-f_{xy}^2, f_x^2f_{yy} - 2f_xf_yf_{xy} + f_{xx}f_y^2\right).
\end{equation*}

\subsubsection{Example}
Consider as our image, the function $f(x, y) = 0.5 - 0.2x - 0.3y - 0.05x^2 +
0.03xy + 0.04y^2 + 0.03y^3 +0.001x^3 - 0.0015x^2y + 0.002xy^2 - 0.0005y^3$
defined on $[0, 1]\times[0, 1]$.
\todo[inline]{remake figure with modified matrix. Also modify function so it has some features like a critical point}
We will subject our image to a special affine transformation:
\begin{equation*}
\mathbf{x} \mapsto \frac{1}{1.1}\begin{bmatrix} 0.6875 & 1.1 \\ -0.6 & 0.8 \end{bmatrix}\mathbf{x} + \begin{bmatrix} 0.1 \\ -0.2 \end{bmatrix}
\end{equation*}

@@

\subsection{The Euclidean Group E(2)}
An element of the Euclidean group $E(2)$ maps an element $(x, y)$ of
$\mathbb{R}^2$ to $(\bar{x}, \bar{y})$ as:
\begin{equation*}
  \begin{aligned}
    \bar{x} &=  x\varepsilon\cos\theta  - y\varepsilon\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
where $\varepsilon \in \{-1, 1\}$ and the remaining parameters are the same
as for $SE(2)$. If $\varepsilon = 1$ it is a rigid transformation, and if
$\varepsilon = -1$ it contains a reflection.
